Grade Retention and Mixed Race

### 2022-12-06 products/paper/main.qmd
Final copy edit

By examining this outcome, I am able to overcome two important limitations that have hampered prior work on the life chances of biracial individuals. First, using self identification of biracial respondents to analyze outcomes is problematic. Research on the identification and classification of biracial individuals indicates that individuals whose parents belong to different races do not always consistently identify with both races [@doyle_are_2007; @liebler_america_2017; @gullickson_racial_2019]. Furthermore, social status and social mobility can predict racial identification generally [@saperstein_racial_2012; @saperstein_mulatto_2013] and social status is related to the identification choices of biracial individuals specifically [@roth_end_2005a; @davenport_role_2016]. Thus, differences in outcomes among biracial individuals may in part produce differences in identification, rather than the other way around.

@@

When children reside with both biological parents, biracial individuals can be identified in large scale data sources such as the Census. Researchers have previously utilized this feature to examine the classification choices made on behalf of these children [@xie_racial_1997; @roth_end_2005a]. Analyzing the outcomes for such children is more difficult, for the simple reason that they do not have many measurable outcomes yet.

However, one important outcome, grade retention, is measurable and important even for young children. Being held back in school is a relatively common occurrence that has consequences for later educational transitions [@jimerson_metaanalysis_2001] and is disproportionately applied by race [@warren_patterns_2014]. Grade retention is arguably one of the earliest stratifiers of life chances that an individual will experience. In this article, I use data from the American Community Survey to examine the likelihood of being held back in school for biracial children in comparison to their monoracial peers. Pooling data from 2010 to 2019, I am able to draw a large enough sample to make reliable statistical inferences for most biracial groups, including those with two non-white parents.

The results provide a window into the placement of biracial children within an existing system of racial inequality in our educational systems. Given the large sample size I am able to draw, I am able to estimate relatively precise estimates for most biracial groups. In contrast to prevailing hypothesis predicting either a pattern of hypodescent or an in-between position for biracial individuals, I find a consistent pattern in which the risk of grade retention is similar to that of the lower risk constituent monoracial group. This result holds even after adjusting for differences in resources across respondents, despite the fact that most biracial groups are positively selected by parental resources relative to their monoracial group. I discuss the implications for these findings in the conclusions.

@@

I distinguish three mechanisms that help us understand where biracials might be positioned within a system of racial inequality. First, I consider arguments for the racial hierarchy that focus on the effects of racial discrimination at both an individual and systemic level. Second, I consider arguments regarding different distributions of resources across racial groups. Third, I consider arguments regarding how the nature of hybridity itself may affect outcomes.

@@

Discrimination and bias by race play an important role in overall levels of racial inequality [@quillian_new_2006; @pager_sociology_2008]. Discrimination need not be consciously recognized by actors, but may instead reflect cognitive biases and schema that individuals unknowningly rely upon and reinforce in social interactions [@quillian_new_2006]. Such discrimination will be most consequential when performed by important gatekeepers of opportunity such as teachers, health care workers, and employers. Discrimination need not just occur on an individual basis, but can also be systemic in nature when biases are built into and reinforced by organizations and institutional systems [@reskin_race_2012]. In our current "colorblind" regime [@bonilla-silva_racism_2006], such systems may be formally race neutral but produce disparate impacts by race [@pager_sociology_2008].

@@

The strongest historical precedent for understanding such classification in the US is the "one drop rule" norm concerning Black ancestry. According to the one-drop rule, individuals of known Black ancestry are considered exclusively Black, regardless of other ancestries [@davis_who_1991]. @davis_who_1991 generalizes this practice into the concept of *hypodescent* in which individuals with mixed racial ancestry will be classified as members of the lower status group. By comparison a pattern of *hyperdescent* implies that individuals of mixed racial ancestry will be classified as members of the higher status group. This pattern of hyperdescent has historically governed the experiences of individuals with American Indian ancestry [@snipp_american_1989; @wolfe_land_2001; @gullickson_choosing_2011].

@@

The development of these particular norms may also reflect the historical regime in which they were developed [@gullickson_choosing_2011; @iverson_regimes_2022]. In this case, neither the hypodescent/hyperdescent or dominance paradigm may help us to fully understand the experiences of those of Latino and Asian ancestry due to more recent processes of racialization. Second, the historical norms for all groups may also be breaking up or transforming in the post-Civil Rights era.

If outcomes for biracial individuals all closer to one of their constituent groups, then we may have stronger evidence about such norms of classification. However, what if those outcomes fall in-between the two groups? How might classification and discrimination produce such an in-between status? There are several possibilities here.

@@

Second, even when "correctly" classified, greater physical ambiguity may allow biracial individuals to avoid the worst instances of bias and discrimination because they do not conform to the ideal type for the group facing the most discrimination. In other words, biracial people's experience of discrimination may be "softened" because their ambiguous physical appearance provokes less racial antipathy. The US in fact already has a long history of this kind of "softening." Because of the one-drop rule, Black individuals vary substantially in skin tone and those individuals with a lighter skin tone have better outcomes [@keith_skin_1991; @hughes_significance_1990; @monk_skin_2014] and are viewed more favorably by both Blacks and Whites [@maddox_cognitive_2002a]. Thus, light skin moderates the stigma of Blackness even among monoracial individuals.

@@

How do these differential distributions of resources affect biracial individuals? A naive argument would be that because biracial individuals have parents from two different racial groups, they will on average inherit parental resources roughly halfway in-between their two constituent groups [@chew_american_1989]. Prior research looking at biracial children in the Census data has found some evidence to support this argument [@chew_american_1989]. In such a case, the resources of biracial individuals would predict outcomes in-between the two constituent groups, barring other mechanisms.

@@

Regardless of its overall prevalence, the net effect of status exchange on resources is complex because status exchange positively selects one parent and negatively selects the other partner. Therefore the two effects may somewhat cancel each other out in terms of the overall distribution of resources across parents. Complicating this issue further, research has shown that the parents of biracial children may differ from parents of monoracial children in the magnitude and types of resources that they allocate to children [@cheng_constraints_2007], even among parents of similar resources.

Regardless of the underlying complexity of this selection process, delineating its role as opposed to other mechanisms is key to understanding the overall process that sorts biracial individuals within the racial hierarchy. Although this approach is sometimes characterized as pitting race versus class and minimizing the importance of race, that is not case here. Instead, what this delineation allows us to do is understand how race affects the outcomes of biracial individuals in a multigenerational process.

@@

Instead of self-identification, researchers have generally relied upon survey data in which the racial identification of a respondent's biological parents are provided in some form or another. @campbell_multiracial_2009 used data from the National Longitudinal Survey of Adolescent to Adult Health (Add Health) to examine high school GPA, advanced placement in math, and four-year college enrollment among biracial and monoracial individuals identified by co-resident biological parents' races. @herman_blackwhiteother_2009 examines differences in GPA using data on high school students in California and Wisconsin in which the students reported their parents' races. Both of these studies find some support for an "in-between" status for many biracial groups, but are both hampered by a lack of statistical power due to small sample sizes. @kao_racial_1999 used data from the National Educational Longitudinal Study of 1988 to examine mathematics achievement scores and GPA for biracial students with Black and Asian parentage and finds that biracial Blacks have outcomes similar to monoracial Blacks, while biracial Asians have outcomes more similar to monoracial Whites. This study, however, is hampered by the fact that biracial students are identified by a discrepancy between their reported race and one parent's reported race, rather than by both biological parent's races.

Results in prior work tend to be imprecise and inconclusive, due to the sample size for most biracial groups. An alternative approach is to use large scale Census or American Community Survey (ACS) data to identify biracial children by the race of the biological parents in their household. This approach can generate much larger samples of biracial individuals, although it does exclude examination of children living in non-nuclear households. Furthermore, researchers then have a limited range of outcomes to explore because biracial individuals identified by this method are by definition children. This approach has been used previously to examine the likelihood of living in poverty [@bratter_multiracial_2013; @bratter_poverty_2013] and residential segregation [@ellis_agents_2012]. These studies provide valuable information on biracial childrens' lived experience, but do not directly measure their own outcomes.

However, one important educational outcome is partially identifiable for children in Census and ACS data - grade retention. I now turn to a closer examination of this particular outcome.

@@

Data for this analysis come from the American Community Survey (ACS), an annual 1-in-100 survey of the US population, conducted by the Census Bureau. To increase sample size for smaller populations of biracial respondents, I pool ACS data for a full decade from 2010-2019. While pooling data helps to increase sample size, it may generate biases if the likelihood of grade retention is changing over time in complex ways. I address this issue in the modeling techniques discussed below. All data were extracted from the Minnesota Population Center's Integrated Public Use Microdata Series [@ruggles_ipums_2020].

@@

The race of children is calculated by a cross-tabulation of parent's races. A full cross-tabulation of the race/Hispanicity of each parent would produce far too many combined categories to be analytically useful. Therefore, I collapse each parent's race into the categories of White, Black, Asian, Latino, and Indigenous. These categories form the "ethnoracial pentagon" that is commonly used in popular practice and government tabulation to identify race in the US [@hollinger_postethnic_1995]. Parents are identified as Latino based on their response to the Hispanicity question, regardless of their response to the race question. Parents are identified as Indigenous if they responded as either American Indian/Alaska Native or as Pacific Islander. Because the goal of my analysis is to examine specifically the outcomes of "first-generation" biracial children, I exclude cases where at least one parent identified with multiple races.

@@

```{r}
#| label: tbl-descriptive
#| tbl-cap: Descriptive survey-weighted statistics by race. Shading indicates biracial group. Groups ordered by sample size.
tbl_summary <- acs_design %>%
  group_by(race) %>%
  summarize(n=n(), 
            held_back=percent(as.numeric(survey_mean(below_exp_grade, 
                                                     vartype=NULL)), 
                              accuracy=0.01),
            mean_age=round(survey_mean(age, vartype=NULL), 1)) %>%
  arrange(desc(n)) %>%
  mutate(n=comma(n))

@@

tbl_summary %>%
  kbl(booktabs = TRUE,
      linesep = NULL,
      align = c("l",rep("r", ncol(tbl_summary)-1)),
      col.names=c("Race",
                  "Sample size",
                  "Clearly behind expected grade", 
                  "Mean age" )) %>%
  kable_styling() %>%
  row_spec(index_biracial, background = "#D3D3D3")
```

@@

The ACS data do not provide the ability to identify or limit survey timing, so I instead use a related measure of whether a given student is clearly behind expected grade (CBEG). A student is considered CBEG if their age is higher than either of the expected ages for a student of that grade. This measure will underestimate overall grade retention because it will miss students who have been retained but not yet had a birthday in their current grade that would place them CBEG. However, the goal of this study is not to estimate grade retention accurately, but rather to understand racial differences in grade retention. Because this bias is largely a function of when students have birthdays, it should be more or less random with regard to sociodemographic characteristics.[^1] This approach is equivalent to the method used by @rosenfeld_nontraditional_2010 on 2000 Census data, but more detailed information on current grade in recent ACS data allows for more precise estimation. @tbl-descriptive shows the percent of each monoracial and biracial group that is CBEG.

[^1]: Biracial respondents are slightly younger than monoracial respondents in the sample as a whole because interracial unions are becoming more common over time which may contribute to a slight downward bias in their risk of grade retention.

My goal is to understand the risk of being CBEG for each biracial group relative to its constituent monoracial groups (e.g. a Black/White student compared to White and Black students). Throughout this article, I use a visual approach to illustrate this placement. As an example, I display the percent of students clearly behind expected grade for White, Black, and Black/White students in @fig-example. I display confidence bands/bars around all three point estimates, but these confidence ranges require some explanation as they are not standard 95% confidence intervals. My goal is to determine whether the biracial group's point estimate is statistically distinguishable from a given monoracial constituent group's point estimate by non-overlap of confidence intervals. Overlap in 95% confidence intervals is not equivalent to failing a hypothesis test of difference at $p<0.05$ and when used in this way, such assessments lead to far more stringent tests [@knol_mis_2011]. When standard errors between groups are equal, overlap in 83.4% confidence intervals will achieve this goal. However, in cases where standard errors vary substantially across estimates, the calculation of the appropriate confidence interval is more complex and varies by each pairwise comparison.[^2]

@@

I begin by showing the relative risk of being CBEG among monoracial respondents across models. Table 2 reports average marginal effects on the probability of being CBEG for each monoracial minority group in comparison to White students. Understanding these differences helps to clarify the potential for how biracial respondents might be positioned between monoracial groups.

@@

@tbl-resources shows the mean values of the resource variables across monoracial and biracial groups.[^7] Because of the number of variables and groups involved, this table can be overwhelming to read. Therefore, I also provide a summation of these resource differentials by racial group by calculating a counterfactual probability of CBEG for each group based on their observed distribution of resource characteristics. This counterfactual probability is calculated by estimating the average predicted probability for each group from the fullest model (Model 4 of Table 2) when all non-resource variables, including race, are held at their mean.[^8]

[^7]: For compactness, I combine the parent foreign-born and English proficiency questions into whether either parent is foreign-born or speak English. In the models, I use separate variables for mothers and fathers. Tables in the supplementary materials provide the full breakdown of these variables by racial group.

@@

My particular interest, however, is in how each biracial group compares to its constituent monoracial groups. For example, while Indigenous/Latino students have the highest counterfactual risk of being CBEG among biracial students, their risk is still lower than their two constituent groups of Indigenous (4.5%) and Latino (5.3%) students, owing in large part to somewhat higher family income, greater parental educational attainment, and greater acculturation. These results suggest positive selection into Indigenous/Latino interracial unions.

@@

The Black/White case stands out as an outlier. Black/White students have similar resources as Black students and both groups have fewer resources than White students. This is observable in @tbl-resources across family income, home ownership, and parental educational attainment. Unlike most of these other groups, there does not seem to be substantial positive class selectivity into Black/White interracial unions. The effect of controlling for these resources will be complex and different across biracial groups. For those biracial groups with resources similar to their more advantaged constituent group, we should see that controlling for these resources should reduce their advantage relative to the less advantaged monoracial group. For the Black/White case, on the other hand, we would expect that controlling for resource differentials should eliminate some of their disadvantage relative to Whites. For the remaining "in-between" groups, it is more difficult to make a priori predictions.

@@

```{r fig-black-models}
#| label: fig-black-models
#| fig-pos: p
#| fig-cap: Probability of being clearly behind expected grade for biracial respondents with one black parent, in comparison to their monoracial comparison groups. . Non-overlap in color corresponding confidence bands indicates statistically significant difference at the 5% level, Baseline model includes year, grade, location, and state fixed effects. Full models include control variables for nativity, English proficiency, income, education, home ownership, and marital status of parents.
#| fig-height: 7.3
coef_table %>%
  filter(model=="Baseline model" | model=="Full model") %>%
  filter(str_detect(mrace, "Black")) %>%
  ggplot(aes(x=term, y=estimate, color=colors))+
  geom_rect(aes(xmax=Inf, xmin=-Inf, ymin=lower, ymax=upper, fill=colors), 
            color=NA, alpha=0.5)+
  geom_linerange(data=subset(ci_df, 
                             (model=="Baseline model" |  model=="Full model") &
                               str_detect(mrace, "Black")), 
                 position=position_dodge2(width = 0.1), size=1, alpha=0.5,
                 aes(ymin=lower, ymax=upper, color=colors))+
  geom_point()+
  coord_flip()+
  facet_grid(mrace~model, scales="free_y")+
  theme_bw()+
  theme(legend.position = "none", panel.grid = element_blank())+
  scale_color_manual(values=color_choices)+
  scale_fill_manual(values=color_choices)+
  scale_y_continuous(labels = scales::percent)+
  labs(x=NULL,
       y="average predicted probability of being clearly behind expected grade")
```

@@

Regardless of variation in biracial students' parental resources, accounting for these resources differentials does not substantially change the overall result that biracial students' risk is more similar to the constituent monoracial group with lower risk. The one exception to this finding is for the Latino/Asian case where the much lower risk of these students compared to Latino and Asian students was entirely driven by resource differentials.

The relative risk for biracial individuals is sometimes incorrectly treated as an indication of a positive future direction for the US in terms of ameliorating racial inequality. The improved prospects for biracial individuals do nothing to ameliorate the often strong and persistent inequalities between monoracial groups. The growth and relative success of biracial populations may isolate remaining members of monoracial minority groups as much as it assimilates mixed race individuals. The real question at stake is how existing divides and identities may be restructured in the context of a growing biracial population. This research suggests the potential for a growing divide between these mixed race populations and the monoracial minority populations from which they derive at least part of their ancestry.

@@

Although sample sizes are big generally, they remain quite small for some groups of non-white biracial respondents, limiting our understanding of the outcomes for those groups. In particular, results for non-white part-Indigenous populations are quite imprecise. The results interestingly and counterintuitively suggest that part-Indigenous populations are the least likely to be in the position expected by hyperdescent, but due to low sample sizes, that finding is highly tentative.

Wile grade retention is an important early life outcome, it is far from the only one that these individuals will face in their life. Although, we often observe similar racial inequalities across a variety of outcomes, knowing the result for one outcome does not perfectly predict other outcomes. As biracial individuals age and face later outcomes in the educational system, the labor market, and elsewhere, their experiences may differ from what I observe here. Thus, collecting more detailed information on these outcomes in future studies remains important.

>>>>

By examining this outcome, I am able to overcome two important limitations that have hampered prior work on the life chances of biracial individuals. First, using self identification of biracial respondents to analyze outcomes is problematic. Prior research indicates that individuals whose parents belong to different races do not always consistently identify with both races [@doyle_are_2007; @liebler_america_2017; @gullickson_racial_2019]. Furthermore, social status and social mobility can predict racial identification generally [@saperstein_racial_2012; @saperstein_mulatto_2013] and social status is related to the identification choices of biracial individuals specifically [@roth_end_2005a; @davenport_role_2016]. Thus, differences in outcomes among biracial individuals may in part produce differences in identification, rather than the other way around.

@@

When children reside with both biological parents, biracial individuals can be identified in large scale data sources such as the Census. Researchers have previously utilized this feature to examine the classification choices made on behalf of these children [@xie_racial_1997; @roth_end_2005a]. Analyzing outcomes for such children is more difficult, for the simple reason that they do not have many measurable outcomes yet. However, one important outcome, grade retention, is measurable and important even for young children. Being held back in school is a relatively common occurrence that has consequences for later educational transitions [@jimerson_metaanalysis_2001] and is disproportionately applied by race [@warren_patterns_2014]. Grade retention is arguably one of the earliest stratifiers of life chances that an individual will experience.

In this article, I use data from the American Community Survey to examine the likelihood of being held back in school for biracial children in comparison to their monoracial peers. Pooling data from 2010 to 2019, I am able to draw a large enough sample to make reliable statistical inferences for most biracial groups, including those with two non-white parents. The results provide a window into the placement of biracial children within an existing system of racial inequality. Given the large sample size I am able to draw, I am able to estimate relatively precise estimates for most biracial groups. In contrast to prevailing hypotheses that predict either a pattern of hypodescent or an in-between position for biracial individuals, I find a consistent pattern in which the risk of grade retention is similar to that of the lower risk constituent monoracial group. This result holds even after adjusting for differences in resources across respondents, despite the fact that most biracial groups are positively selected by parental resources relative to their monoracial group. I discuss the implications for these findings in the conclusions.

@@

I distinguish three mechanisms that help us understand where biracial individuals might be positioned among these possibilities. First, I consider arguments that focus on the effects of racial discrimination at both an individual and systemic level. Second, I consider arguments regarding different distributions of resources across racial groups. Third, I consider arguments regarding how the nature of hybridity itself may affect outcomes.

@@

Discrimination and bias by race play an important role in overall levels of racial inequality [@quillian_new_2006; @pager_sociology_2008]. Discrimination need not be consciously recognized by actors, but may instead reflect cognitive biases and schema that individuals unknowingly rely upon and reinforce in social interactions [@quillian_new_2006]. Such discrimination will be most consequential when performed by important gatekeepers of opportunity such as teachers, health care workers, and employers. Discrimination need not just occur on an individual basis, but can also be systemic in nature when biases are built into and reinforced by organizations and institutional systems [@reskin_race_2012]. In our current "colorblind" regime [@bonilla-silva_racism_2006], such systems may be formally race neutral but produce disparate impacts by race [@pager_sociology_2008].

@@

The strongest historical precedent for understanding such classification in the US is the "one drop rule" norm governing classification as Black. According to the one-drop rule, individuals of known Black ancestry are considered exclusively Black, regardless of other ancestries [@davis_who_1991]. @davis_who_1991 generalizes this practice into the concept of *hypodescent* in which individuals with mixed racial ancestry will be classified as members of the lower status group. By comparison a pattern of *hyperdescent* implies that individuals of mixed racial ancestry will be classified as members of the higher status group. This pattern of hyperdescent has historically governed the experiences of individuals with American Indian ancestry [@snipp_american_1989; @wolfe_land_2001; @gullickson_choosing_2011].

@@

The development of these particular norms may also reflect the historical regime in which they were developed [@gullickson_choosing_2011; @iverson_regimes_2022]. In this case, neither the hypodescent/hyperdescent or dominance paradigm may help us to fully understand the experiences of those of Latino and Asian ancestry due to more recent processes of racialization. Furthermore, the historical norms for all groups may be weakening or transforming in the post-Civil Rights era.

If outcomes for biracial individuals all closer to one of their constituent groups, then we may have stronger evidence about such norms of classification. However, what if those outcomes fall in-between the two groups? How might classification and discrimination produce such an in-between status? Several scenarios might produce such an in-between status.

@@

Second, even when "correctly" classified, greater physical ambiguity may allow biracial individuals to avoid the worst instances of bias and discrimination because they do not conform to the ideal type for the group facing the most discrimination. In other words, biracial people's experience of discrimination may be "softened" because their ambiguous physical appearance provokes less racial antipathy. The US already has a long history of this kind of softening. Because of the one-drop rule, Black individuals vary substantially in skin tone and those individuals with a lighter skin tone have better outcomes [@keith_skin_1991; @hughes_significance_1990; @monk_skin_2014] and are viewed more favorably by both Blacks and Whites [@maddox_cognitive_2002a]. Thus, light skin moderates the stigma of Blackness even among monoracial individuals.

@@

How do these differential distributions of resources affect biracial individuals? A naive argument would be that because biracial individuals have parents from two different racial groups, they will on average inherit parental resources roughly halfway between their two constituent groups [@chew_american_1989]. Prior research looking at biracial children in the Census data has found some evidence to support this argument [@chew_american_1989]. In such a case, the resources of biracial individuals would predict outcomes in-between the two constituent groups, barring other mechanisms.

@@

Regardless of its overall prevalence, the net effect of status exchange on resources is complex because status exchange positively selects one partner and negatively selects the other partner. Therefore the two effects may somewhat cancel each other out in terms of the overall distribution of resources across parents. Complicating this issue further, research has shown that the parents of biracial children may differ from parents of monoracial children in the magnitude and types of resources that they allocate to children [@cheng_constraints_2007], even among parents of similar resources.

Regardless of the underlying complexity of this selection process, delineating its role as opposed to other mechanisms is key to understanding the overall process that sorts biracial individuals within the racial hierarchy. Although this approach is sometimes characterized as pitting race versus class and minimizing the importance of race, that is not the case here. Instead, this delineation allows us to understand how race affects the outcomes of biracial individuals in a multigenerational process.

@@

Instead of self-identification, researchers have generally relied upon survey data in which the racial identification of a respondent's biological parents are provided in some form or another. @campbell_multiracial_2009 used data from the National Longitudinal Survey of Adolescent to Adult Health (Add Health) to examine high school GPA, advanced placement in math, and four-year college enrollment among biracial and monoracial individuals identified by co-resident biological parents' races. @herman_blackwhiteother_2009 examines differences in GPA using data on high school students in California and Wisconsin in which the students reported their parents' races. Both of these studies find some support for an "in-between" status for many biracial groups, but are both hampered by a lack of statistical power due to small sample sizes. @kao_racial_1999 used data from the National Educational Longitudinal Study of 1988 to examine mathematics achievement scores and GPA for biracial students with Black and Asian parentage and finds that biracial Black students have outcomes similar to monoracial Black students, while biracial Asian students have outcomes more similar to monoracial White students. This study, however, is hampered by the fact that biracial students are identified by a discrepancy between their reported race and one parent's reported race, rather than by both biological parents' races.

Results in prior work tend to be imprecise and inconclusive, due to the sample size for most biracial groups. An alternative approach is to use large scale Census or American Community Survey (ACS) data to identify biracial children by the race of the biological parents in their household. This approach can generate much larger samples of biracial individuals, although it does exclude examination of children living in non-nuclear households. However, researchers then have a limited range of outcomes to explore because biracial individuals identified by this method are by definition children. This approach has been used previously to examine the likelihood of living in poverty [@bratter_multiracial_2013; @bratter_poverty_2013] and residential segregation [@ellis_agents_2012]. These studies provide valuable information on biracial childrens' lived experience, but do not directly measure their own outcomes. However, one important educational outcome is partially identifiable for children in Census and ACS data - grade retention. I now turn to a closer examination of this particular outcome.

@@

Data for this analysis come from the American Community Survey (ACS), an annual 1-in-100 survey of the US population, conducted by the Census Bureau. To increase sample size for smaller populations of biracial respondents, I pool ACS data for a full decade from 2010-2019. All data were extracted from the Minnesota Population Center's Integrated Public Use Microdata Series [@ruggles_ipums_2020].

@@

The race of children is calculated by a cross-tabulation of parents' races. A full cross-tabulation of the race/Hispanicity of each parent would produce far too many combined categories to be analytically useful. Therefore, I collapse each parent's race into the categories of White, Black, Asian, Latino, and Indigenous. These categories form the "ethnoracial pentagon" that is commonly used in popular practice and government tabulation to identify race in the US [@hollinger_postethnic_1995]. Parents are identified as Latino based on their response to the Hispanicity question, regardless of their response to the race question. Parents are identified as Indigenous if they responded as either American Indian/Alaska Native or as Pacific Islander. Because the goal of my analysis is to examine specifically the outcomes of "first-generation" biracial children, I exclude cases where at least one parent identified with multiple races.

@@

```{r}
#| label: tbl-descriptive
#| tbl-cap: Descriptive survey-weighted statistics by race. Shading indicates biracial group. Groups ordered by sample size.
tbl_summary <- acs_design %>%
  group_by(race) %>%
  summarize(n=n(), 
            held_back=percent(as.numeric(survey_mean(below_exp_grade, 
                                                     vartype=NULL)), 
                              accuracy=0.01)) %>%
  arrange(desc(n)) %>%
  mutate(n=comma(n))

@@

tbl_summary %>%
  kbl(booktabs = TRUE,
      linesep = NULL,
      align = c("l",rep("r", ncol(tbl_summary)-1)),
      col.names=c("Race",
                  "Sample size",
                  "Clearly behind expected grade")) %>%
  kable_styling() %>%
  row_spec(index_biracial, background = "#D3D3D3")
```

@@

The ACS data do not provide the ability to identify or limit survey timing, so I instead use a related measure of whether a given student is clearly behind expected grade (CBEG). A student is considered CBEG if their age is higher than either of the expected ages for a student of that grade. This measure will underestimate overall grade retention because it will miss students who have been retained but not yet had a birthday in their current grade that would place them CBEG. However, the goal of this study is not to estimate grade retention accurately, but rather to understand racial differences in grade retention. Because this bias is largely a function of when students have birthdays, it should be more or less random with regard to sociodemographic characteristics. This approach is equivalent to the method used by @rosenfeld_nontraditional_2010 on 2000 Census data, but more detailed information on current grade in recent ACS data allows for more precise estimation. @tbl-descriptive shows the percent of each monoracial and biracial group that is CBEG.

My goal is to understand the risk of being CBEG for each biracial group relative to its constituent monoracial groups (e.g. a Black/White student compared to White and Black students). Throughout this article, I use a visual approach to illustrate this placement. As an example, I display the percent of White, Black, and Black/White students clearly behind expected grade in @fig-example. I display confidence bands/bars around all three point estimates, but these confidence ranges require some explanation as they are not standard 95% confidence intervals. My goal is to determine whether the biracial group's point estimate is statistically distinguishable from a given monoracial constituent group's point estimate by non-overlap of confidence intervals. Overlap in 95% confidence intervals is not equivalent to failing a hypothesis test of difference at $p<0.05$ and when used in this way, such assessments lead to far more stringent tests [@knol_mis_2011]. When standard errors between groups are equal, overlap in 83.4% confidence intervals will achieve this goal. However, in cases where standard errors vary substantially across estimates, the calculation of the appropriate confidence interval is more complex and varies by each pairwise comparison.[^2]

@@

I begin by showing the relative risk of being CBEG among monoracial respondents across models.Understanding these differences helps to clarify the potential for how biracial respondents might be positioned between monoracial groups. Table 2 reports average marginal effects on the probability of being CBEG for each monoracial minority group in comparison to White students.

@@

@tbl-resources shows the mean values of the resource variables across monoracial and biracial groups.[^7] I also calculate a counterfactual probability of CBEG for each racial group based solely on their observed distribution of resources. Differences in this counterfactual probability provides a summary measure of the resource differentials between groups. This counterfactual probability is calculated by estimating the average predicted probability for each group from the fullest model (Model 4 of Table 2) when all non-resource variables, including race, are held at their mean.[^8]

[^7]: For compactness, I combine the parent foreign-born and English proficiency questions into whether either parent is foreign-born or speaks English. In the models, I use separate variables for mothers and fathers. Tables in the supplementary materials provide the full breakdown of these variables by racial group.

@@

I am particularly interested in how each biracial group compares to its constituent monoracial groups. For example, while Indigenous/Latino students have the highest counterfactual risk of being CBEG among biracial students, their risk is still lower than their two constituent groups of Indigenous (4.5%) and Latino (5.3%) students, owing in large part to somewhat higher family income, greater parental educational attainment, and greater acculturation. These results suggest positive selection into Indigenous/Latino interracial unions.

@@

The Black/White case stands out as an outlier. Black/White students have similar resources as Black students and both groups have fewer resources than White students. This is observable in @tbl-resources across family income, home ownership, and parental educational attainment. Unlike most of these other groups, there does not seem to be substantial positive class selectivity into Black/White interracial unions.

The effect of controlling for these resources will be complex and different across biracial groups. For those biracial groups with resources similar to their more advantaged constituent group, controlling for these resources should reduce their advantage relative to the less advantaged monoracial group. For the Black/White case, on the other hand, controlling for resource differentials should eliminate some of their disadvantage relative to Whites. For the remaining "in-between" groups, it is more difficult to make a priori predictions.

@@

```{r fig-black-models}
#| label: fig-black-models
#| fig-pos: p
#| fig-cap: Probability of being clearly behind expected grade for biracial respondents with one black parent, in comparison to their monoracial comparison groups. Non-overlap in color corresponding confidence bands indicates statistically significant difference at the 5% level, Baseline model includes year, grade, location, and state fixed effects. Full models include control variables for nativity, English proficiency, income, education, home ownership, and marital status of parents.
#| fig-height: 7.3
coef_table %>%
  filter(model=="Baseline model" | model=="Full model") %>%
  filter(str_detect(mrace, "Black")) %>%
  ggplot(aes(x=term, y=estimate, color=colors))+
  geom_rect(aes(xmax=Inf, xmin=-Inf, ymin=lower, ymax=upper, fill=colors), 
            color=NA, alpha=0.5)+
  geom_linerange(data=subset(ci_df, 
                             (model=="Baseline model" |  model=="Full model") &
                               str_detect(mrace, "Black")), 
                 position=position_dodge2(width = 0.1), size=1, alpha=0.5,
                 aes(ymin=lower, ymax=upper, color=colors))+
  geom_point()+
  coord_flip()+
  facet_grid(mrace~model, scales="free_y")+
  theme_bw()+
  theme(legend.position = "none", panel.grid = element_blank())+
  scale_color_manual(values=color_choices)+
  scale_fill_manual(values=color_choices)+
  scale_y_continuous(labels = scales::percent)+
  labs(x=NULL,
       y="average predicted probability of being clearly behind expected grade")
```

@@

Regardless of variation in biracial students' parental resources, accounting for these resource differentials does not substantially change the overall result that biracial students' risk is more similar to the constituent monoracial group with lower risk. The one exception to this finding is for the Latino/Asian case where the much lower risk of these students compared to Latino and Asian students was entirely driven by resource differentials.

Low relative risk for biracial individuals is sometimes treated as an indication of a positive future direction for the US in terms of ameliorating racial inequality. However, the improved prospects for biracial individuals do nothing to ameliorate the often strong and persistent inequalities between monoracial groups. The growth and relative success of biracial populations may isolate remaining members of monoracial minority groups as much as it assimilates mixed race individuals. The real question at stake is how existing divides and identities may be restructured in the context of a growing biracial population. This research suggests the potential for a growing divide between these mixed race populations and the most disadvantaged monoracial minority populations from which they derive at least part of their ancestry.

@@

Sample sizes for some non-white biracial respondents remain small, limiting our understanding of the outcomes for those groups. In particular, results for non-white part-Indigenous populations are quite imprecise. The results interestingly and counterintuitively suggest that part-Indigenous populations are the least likely to be in the position expected by hyperdescent, but due to low sample sizes, that finding is highly tentative.

Wile grade retention is an important early life outcome, it is far from the only one that these individuals will face in their life. Although, we often observe similar racial inequalities across a variety of outcomes, knowing the result for one outcome does not perfectly predict other outcomes. As biracial individuals age and face later outcomes in the educational system, the labor market, and elsewhere, their experiences may differ from what I observe here. Thus, collecting more detailed information on these outcomes in future studies remains critically important.
